Discussion Draft

### 2020-03-24 doc/Discussion.rmd
re-wrote outline of chromatin org section

# Explanations and Model testing

@@

This is how we are going to test/ synthesize / fit our results into the literuature and test hypothesis for which models best explain our results. In our analysis / synthesis of results, we decded to systematically compared models which have been proposed to explain our results (both for the sex dfferences and the faster male evolution of recombination rates). We attempted to test the hypotheses of if our choosen models could produce the main / general patterns oberved in this study (Table X)

We generally focus on models for the evolution of heterochiasmy since that is the primary data/set of results in this study. We search for a parsimonius explanation /model that fits both the results from sex differences and the faster male evolution. But we also test the male specific traits.

@@

**Conserved heterochiasmy paterns**

Below are the main traits from our results: more between cell variance in females, chromatin compaction (with consequence for interference strength), and broad scale landscape differences.

@@

```{r chromatin.organization, include=FALSE}

#1. chromatin organization, female shorter loops, longer axis

#2. A consequence of chromatin organization (loop size) is difference in interference strength in (MB / cM)

#seen in humans, mice (signal comes through even with cell to cell variation in chromatin compaction / organization)
#

#exceptions to female pattern (fish birds), lack of data
# interference strength reversal (At and cattle) -- we predict that the loop size will also reverse, -- but there is a lack of cytological data for both sexes.

#lots of data on genetic maps for both sexes, but lack of cytological data to support

#our results, female longer axis, 
#review caveats -- why we believe the pattern is true

#refs
#BF klecner?
#negative relationship of axis and loop size (Lynn? Gruhn?)
#global re-orgnaiziton of DNA / chromosomes (3D, new Chip-seq pubs)
```

1. basics and expectations for chromatin organization

2. Well supported model for interference is that it is mediated / communicated through the 2D chromatin axis or SC (depending on the timing)

3. consequence for interference (general patterns) is that interference measured as (mb) will be stronger if DNA loops are stronger

The relationship // molecular organization of chromosome -- undergoes global re-orgzniation during meiosis. The basic take away is that eraly stage in meiosis) there is a central axis made up of chromatin and proteins while most of the DNA / chromatin are folded into loops that emenate off of the axis. This structure has been called 'lampbrush' by some (cite).

A pattern that has been long noticed --

Merging of two types of data

-total chromatin structure (axis length) (cytology)

-Interfere measured as cM or DNA base pairs Mb  (the spacing between markers from linkage map)

general pattern for the sexual dimorphsim the relationship between axis length and CO number
-- axis are longer in females than male, suggesting suport more crossovers (more DSBs)
but the density is less in females.

Our results from MSM and PWD show that the sc length compaction is independant (of the direction of heterochiasmy. Two strains have have more crossovers in males, but females still have longer sc axis. _This indecates that the sexual dimorphism in chromosome organization / DNA compaction is a conserved part of recombination landscape_

**lack of difference in the raw measures of IFD, but significant difference in **

The pattern of interference across raw and normalized measures -- also confirms previous results of male having stronger interfere -- due to the longer loops (Petkov) **these results mirror those seen in** [@petkov2007]

< Petkov paper -- states that the sexual dimorphism for interference strength is due to the chromatin structure (meiotic chromosome organization) with the increased interference strength in males being due to longer chromatin loop lengths.>

<same pattern in humans (Gruhn et al 2013)
  (exceptions, birds (fish?) plants?) -- examining this is hard due to chromosome size effecs
  - This also assumes that the DNA loop sizes have uniform size along and between chromosomes>

2. Short axis / longer DNA loops will have stronger interference
This prediction combines the findings of chromatin compaction and interference differences which are intertwined by the transmission of interference along the chromosome axis and the inverse relationship chromatin compaction (axis length and DNA loop sizes).
In physical units (micron) the counting model fits interference well to the 2D linear space of the axis these measures of interference are conserved across sexes in mouse. This is because the counting units are the BASES of DNA loops which is conserved along the 2D length of chromatin axis. When this space is transformed to account for all of the DNA base pairs which are sequestered into the DNA loops above the axis, the strength of interference will vary depending the on underlying chromatin structure. 
The chromatin structure which has longer loops (and shorter axis) will have more DNA bps folded / sequestered into the DNA loops. Thus the strength of interference will be stronger because with each step (count) along the 2D axis, 
more bases pairs are being skipped above the axis.

@@

@@

#revrsed heterochiasmy direction
#faster -- rapid evoluiton in males

#this also goes along with

# interference strength

Some of the traits related to faster male evolution -- distinguish in which models can redict them:

Our results had two notable instances of evolution in the recombination rate (or correlated trait); evolution of precursor number (DSBs) and crossover interference strength. <While SC length and the telomere distances are similar / conserved across the mouse strains>

# Evolved rates of precursors, DSBs

```{r evo.DSB, include=FALSE}

<define CO:NCO decision, tethered-loop DSB formation SPO11 (cite, paigen? deMassy
setup baier study,  explain >

_the propotion of DSB repaired into crossovers hasn't evolved, the total number of DSBs has evolved_  (following the tethered lop model of DSBs formation -- this would suggest that in the high rec strains, assuming no large difference in the Mb size of the chromosomes, 
during chromatin remodeling in prophase, in the high rec strains the chromosomes have longer axis, allowing a higher number but shorter DNA loops (for SPO11 to make the DSBs).

Our results find evolution between a subset of strains -- demonstrating that the high rec strains have significant higher DSB number.

Our results mirror those found in Baier et al [@baier2014]

@@

Baier et al found significant differences in sc legnths for 
The tests in our study - show moderate to low significance in the differences in SC lengths. (this is because we have more noise-can't distinguish individual chromosomes)

our results are similar in several ways
- correlation of final MLH1 and DBS counts, SC measues correlate with the DSB number (even tho the differnces we observe in our study are -- slight / less significant -- we attribute that to chromosome specific noise )

Evolution of recombination rates via the number of precursors -- (has been proposed -- ) (Bairer et al).

@@

#models that predict this pattern?

@@

1. Review results
-The rapid evolution of the genome wide rate in our high rec strains is due to more enrichment of 2CO which have distinct rec landscape from the 2COs in low rec strains.

@@

@@

>>>>

@@

To synthesize our results, we decded to systematically compare models which have been proposed for explaining the evolution of heterochiasmy or meiotic machinery. There are two main types of models: modifers (indirect) modifers or functional (cell physiology based models). We attempted to test the hypotheses (mental test for if the model) (Table X).

@@

@@

@@

```{r chromatin.organization, include=FALSE}
#1. basics and expectations for chromatin organization
#seen in humans, mice (signal comes through even with cell to cell variation in chromatin compaction / organization)

#2. how the model treat this limiting factor

#3. consequence for interference (general patterns) is that interference measured as (mb) will be stronger if DNA loops are stronger
#prediction for general -- pattern
```

0.  Our results from MSM and PWD show that the sc length compaction is independant (of the direction of heterochiasmy. Two strains have have more crossovers in males, but females still have longer sc axis. _This indecates that the sexual dimorphism in chromosome organization / DNA compaction is a conserved part of recombination landscape / heterochiasmy_

1. the Chromosome axis length // chromatin organizatio in meiotisis

according to the tethered loop model of DSB formation -- the 'real estate' for COs are the DNA loops - the density of whch is conserved in the axis and loop formation (later the SC str), so the number of loops is equvilent to the length of the axis. When the total amount of DNA is kept constant as for a given Chromsome, there is a negative relationship between the loop length and axi length. **The chromosomes are converted to a 2D space (the axis length of a given chromosome is the limited factor for total CO number)**   This structure has been called 'lampbrush' by some (cite).

2. Only one of the models we examined predicts sexual dimorphism for axis length.

The modeifer models do not parameterize this since they do not models / are not built to consider single meiotic events.

**COM** predicts that axis length will be larger in oocytes due to the larger volumne -- this would help in the bouquet stage -- may be connected to the 'the search' during homolog pairing.

the **spindle based selection** model does not specific predictions about this aspect -- assumes that the 2D areas for chromosomes are proportaional across stages (during the decondensing stage between pachtyene and dipl? / metaphase) -- the chromosomes condenses uniformly across the length.

3. another thing to point out / regaurding this model -- is the connection to sex difference in interference strength. Our results of conserved within strain raw IFD lengths - yet when the IFDs are controled for total SC length, we observed sexual dimorphism with males having longer IFD^normalized^ (stronger interference), mirror those observed in [@petkov2007].
The model / explaination proposed in that paper is that The chromatin structure which has longer loops (and shorter axis) will have more DNA bps folded / sequestered into the DNA loops. Thus the strength of interference will be stronger because with each step (count) along the 2D axis, more bases pairs are being skipped above the axis. These findings point out two things, This part of heterochiasmy generates testable prediction: First, the interference strength measured in chromsoome axis units (or SC) is generally conserved between sexes. Second the sex with longer axis and shorter loops will have weaker interference compared to the other sex. Or conversely, the sex with stronger linkage map based interference woud have chromatin organization with shorter axis compared to the other sex. There is a plefora of data of sexual dimorphism in linkage map based interference strength indeicating sexual dimorphism, but comparative cytological data is lacking (difference in perparness of some sexes and complications due to idenification of specific chromosomes between cells.)

@@

#tricker to fit the models to this, 
#paradox -- how the broad/general patterns from the sex differences

# while DSBs have evolved between these groups - not integrated into the table for models -- since most models don't integrate this stage of the pathway.

#1.reversed heterochiasmy direction
#faster -- rapid evoluiton in males

#2 interference strength (intro)

#SACE could fit ...

#spindle based selection 
```

- the male specific differences in meiotic traits are a bit trickier to explain using the above models (since they were designed to explain broad scale pattern).

The general patterns observed from the within strain / between sex data --generate predictions for the relationship between chromatin compaction and interference strength. (the predictions are). Based on the general / braod scale patterns, those predictions are not met -- but we argue that our data are not suited / don't have enough power to -- overcome the weaker signal within the nose of chromosome specific and within nucleus correlation in chromatin compaction. (also distingushing the recombination rate evolution at this scale is not the focus of this manuscript.)

Some of the traits related to faster male evolution -- distinguish in which models can redict them: the reverseal of heterochiasmy direction and the correlated evolution of gwRR and interference strength.

**DSB evolved but, don't included as an examined trait, since most evolution based models don't make predictions for this stage of meiosis.**
_the propotion of DSB repaired into crossovers hasn't evolved, the total number of DSBs has evolved_  (following the tethered lop model of DSBs formation -- this would suggest that in the high rec strains, assuming no large difference in the Mb size of the chromosomes, 
during chromatin remodeling in prophase, in the high rec strains the chromosomes have longer axis, allowing a higher number but shorter DNA loops (for SPO11 to make the DSBs). Our results mirror those found in Baier et al [@baier2014]

@@

# revered heterochiasmy

```{r reverse.HetC, echo=FALSE}

models for haploid selection -- this **reversal of the direction of heterochiasmy** would be accompanied by (vast changes in haploid stages or degree of imprinting)

the meiotic drive -- model, females should be predicted to have larger linkage maps/gwRR since they are more exposed / weaker to true meiotic drivers.

A reverselal in heteroschiasmy is not clearly predicted by the sace model (unless with the -- except under the conditions discribed below)

@@

#models that predict this pattern

@@

We had an unexpectedd result of correlated evolution of gwRR with interference strength. This is unexpected beacuse...

-The rapid evolution of the genome wide rate in our high rec strains is due to more enrichment of 2CO which have distinct rec landscape from the 2COs in low rec strains.

@@

**SACE**
a reversal in heterochiasmy direction would make sense (under the sae model) if it maintained 
this isn't an explict prediction, but accordding this this model -- stronger interference (such as we observed) would have the consequence of maintaining larger blocks of chromosomes together.

@@

<Petkov paper -- states that the sexual dimorphism for interference strength is due to the chromatin structure (meiotic chromosome organization) with the increased interference strength in males being due to longer chromatin loop lengths.

same pattern in humans (Gruhn et al 2013)
  (exceptions, birds (fish?) plants?) -- examining this is hard due to chromosome size effecs
  - This also assumes that the DNA loop sizes have uniform size along and between chromosomes

2. Short axis / longer DNA loops will have stronger interference
This prediction combines the findings of chromatin compaction and interference differences which are intertwined by the transmission of interference along the chromosome axis and the inverse relationship chromatin compaction (axis length and DNA loop sizes).
In physical units (micron) the counting model fits interference well to the 2D linear space of the axis these measures of interference are conserved across sexes in mouse. This is because the counting units are the BASES of DNA loops which is conserved along the 2D length of chromatin axis. When this space is transformed to account for all of the DNA base pairs which are sequestered into the DNA loops above the axis, the strength of interference will vary depending the on underlying chromatin structure.>

<Baier et al found significant differences in sc legnths for 
The tests in our study - show moderate to low significance in the differences in SC lengths. (this is because we have more noise-can't distinguish individual chromosomes)

our results are similar in several ways
- correlation of final MLH1 and DBS counts, SC measues correlate with the DSB number (even tho the differnces we observe in our study are -- slight / less significant -- we attribute that to chromosome specific noise )>
